---
title: "Alignment, Clocking, and Macro Patterns of Episodes in the Life Course"
author: "Tim Riffe, Angelo Lorenti, Andrés Castro"
date: "1 July 2022" 
colortheme: "dove"
output: beamer_presentation
header-includes:
  - \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Example questions:

\Large
> - Do disability episodes get shorter or longer with age? And over time?
> - How much of a state expectancy is composed of short vs long episodes?
> - How do parity-specific birth interval distributions vary by completed fertility or in response to birth outcomes?

## 

\Large
\begin{block}{\textbf{Solution}}
We develop a framework (or grammar) of data operations to flexibly derive aggregate patterns from trajectory data.
\end{block}

<br>
\pause

\begin{block}{\textbf{Approach}}
 \textbf{Clocks} are within and between episode timekeeping operations.\\
 \textbf{Alignment} is a time structuring operation.  
\end{block}
<!-- if simulated then there is likely a markov expression waiting to be written -->

## Approach

\Large
\begin{block}{\textbf{Clocks}}
Within episodes of state \textbf{s}, count time \textcolor{red}{steps} or episode  \textcolor{red}{order} up or down, or total episode  \textcolor{red}{duration} conditional on time of episode entry, exit, or neither.
\end{block}

## Approach

\Large
\begin{block}{\textbf{Alignment}}
  \textcolor{red}{left}, \textcolor{red}{right}, \textcolor{red}{center}, etc. on the \textcolor{blue}{first}, \textcolor{blue}{last}, \textcolor{blue}{longest}, \textcolor{blue}{shortest}, \textcolor{blue}{$n^{th}$}, \textcolor{blue}{$n^{th}$ from last} episode of state \textbf{s}. 
\end{block}

## Requisites

\Large 
\begin{block}{\textbf{Trajectory data}}
 A set of either \textcolor{red}{observed} or \textcolor{red}{simulated} time series of \textcolor{blue}{discrete time steps} consisting in  \textcolor{blue}{categories}. 
\end{block}

## Illustrations

\Large
10 lives simulated from Dudel & Myrskylä (2017)
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
if (! "Spells" %in% rownames(installed.packages())){
  remotes::install_github("timriffe/Spells/R/Spells")
}
library(tidyverse)
library(Spells)
library(colorspace)
library(here)
library(reshape2)
library(ggridges)
source(here::here("R","00_load_functions.R"))
#cols   <- c(qualitative_hcl(5, palette = "Dark 3")[c(3,2,4)],NA)
palette_OkabeIto <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
cols             <- c(palette_OkabeIto[c(3,1,2)],NA)
# plot(NULL, xlim=c(0,length(palette_OkabeIto)), ylim=c(0,1), 
#     xlab="", ylab="", xaxt="n", yaxt="n")
# rect(0:(length(palette_OkabeIto)-1), 0, 1:length(palette_OkabeIto), 1, col=palette_OkabeIto)

# background colors for clock measures:
cols2            <- lighten(desaturate(cols,.3),.3)
states           <- c("Employed", "Inactive","Retired","Dead")
yvals            <- 9:0 * 1.1
X                <- readRDS(here("Talks","CED","X.rds"))
XX               <- readRDS(here("Talks","CED","RTraj_clean.rds"))
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Age structured prevalence.

\Large 
\emph{Identity} clock in employment state
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Ones            <- X == "Employed"
Ones[Ones]      <- 1
Ones[Ones == 0] <- NA
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(Ones[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Age structured prevalence.

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Ones     <- XX == "Employed"
Ones[XX == "Dead"] <- NA
PrevEmpl <- rowMeans(Ones, na.rm=TRUE)
```
```{r}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,1), axes = FALSE, xlab = "", ylab = "")
lines(50:100, PrevEmpl, lwd = 2)
axis(1)
text(43,.5,"Prevalence",xpd=TRUE,srt=90)
axis(2, las = 1)
```

## Illustration: Clocks: Duration (unconditional)

```{r}
# unconditional durations
Dur      <- apply(X,2,clock,clock_type="duration",state = "Inactive")
# conditional on age at entry:
# (function not yet wrapped in clock())
DurEntry      <- apply(X,2,spell_dur_conditional,state = "Inactive", entry = TRUE)
DurExit      <- apply(X,2,spell_dur_conditional,state = "Inactive", entry = FALSE)
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(Dur[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Duration \textcolor{red}{conditioned on entry}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(DurEntry[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Duration \textcolor{blue}{conditioned on exit}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(DurExit[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Order \textcolor{red}{Ascending}

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
OrdUp   <- apply(X,2,clock,clock_type="order",state = "Employed",increasing=TRUE, step_size = 1)
OrdDown <- apply(X,2,clock,clock_type="order",state = "Employed",increasing=FALSE, step_size = 1)
```
```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(OrdUp[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Order \textcolor{blue}{Descending}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(OrdDown[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Steps \textcolor{red}{Ascending}

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
StepUp   <- apply(X,2,clock,clock_type="step",state = "Inactive",increasing=TRUE, step_size = 1)
StepDown <- apply(X,2,clock,clock_type="step",state = "Inactive",increasing=FALSE, step_size = 1)
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(floor(StepUp)[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Steps \textcolor{blue}{Descending}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(floor(StepDown)[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: Age = Birth alignment

```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,130), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Alignment: Death

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
XdeathAlign <- apply(X,2,align,state = "Dead",type="left")
```

```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-80,0), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=XdeathAlign[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-80,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-83,6,"Random individual i",xpd=TRUE,srt=90)
legend(-40,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Alignment: \emph{Entry} to \emph{first} retirement
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
XretirefirstAlign <- apply(X,2,align,state = "Retired",type="left",spell = "first")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=XretirefirstAlign[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-33,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: \emph{Exit} from \emph{first} employment
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Xemployexitfirst <- apply(X,2,align,state = "Employed",type="right",spell = "first")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=Xemployexitfirst[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-33,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: \emph{Exit} from \emph{longest} employment
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Xemployexitlongest<- apply(X,2,align,state = "Employed",type="right",spell = "longest")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=Xemployexitlongest[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-33,6,"Random individual i",xpd=TRUE,srt=90)
```

## Aggregation

\Large
\begin{block}{\textbf{Macro patterns}}
Combine clocks and alignment to aggregate (e.g. means, quantiles)
\end{block}

## A Health Application 
```{r statespace, echo = FALSE,warning=FALSE,message=FALSE}
# tools::texi2dvi('statespace.tex',pdf=TRUE,clean=TRUE)
# a<-file.rename(from="statespace.pdf",to="statespace.pdf")
# system(paste("cd",here::here(), "&& inkscape -l statespace.svg statespace.pdf",sep=" "))
knitr::include_graphics("statespace.pdf")
```
```{r, include = FALSE}
A1.1 <- readRDS(here("Data","Application1","A1.1.rds"))
A1.2 <- readRDS(here("Data","Application1","A1.2.rds"))
A1.3 <- readRDS(here("Data","Application1","A1.3.rds"))
```


## Inequality in disability spell duration
```{r, message=FALSE,warning=FALSE, echo = FALSE}
A1.1 %>% 
  group_by(INC_Q, age) %>% 
  summarize(lower = quantile(dur_first_mean,0.025),
            upper = quantile(dur_first_mean,0.975),
            median = median(dur_first_mean),
            .groups = "drop") %>% 
  ggplot(mapping = aes(x = age, 
                       y = median, 
                       color = INC_Q,
                       fill = INC_Q,
                       ymin = lower,
                       ymax = upper)) + 
  geom_ribbon(alpha=.2, color = NA) +
  geom_line(size=1.5) + 
  geom_segment(aes(x=38,y=2,xend=49,yend=2),color = "black") +
  annotate("text", x = 38, y = 1.4, label = "highest 20%") +
  annotate("text", x = 35, y = 2.2, label = "lowest 20%") +
  annotate("text", x = 45, y = 1.4, label = "Women from the highest\nincome quintile reach a mean\ndisability bout duration\nof 2 years 10 years later.", hjust=0)+
  geom_curve(mapping=aes(x=43.5,y=1.65,xend = 45,yend=2),
             color="black",size=.1,curvature=-.6,arrow = arrow(length = unit(0.02, "npc")))+
  xlim(26,65) +
  labs(x = "Age", y = "conditional mean spell duration (years)",
       main = "Mean disability spell duration of spells starting in age x") +
  guides(color = "none", fill = "none")+
  theme(
    axis.title.x = element_text(size = 16), # use 16 for presentations
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.y = element_text(size = 16)) +
  theme_minimal() +
  scale_fill_brewer(palette = "Paired") + 
  scale_color_brewer(palette = "Paired")
```


## How many times have people been disabled? 

```{r, echo = FALSE, warning = FALSE, message = FALSE}
  A1.2 %>% 
  group_by(INC_Q, age) %>% 
  summarize(lower = quantile(order_first_mean,0.025),
            upper = quantile(order_first_mean,0.975),
            median = median(order_first_mean),
            .groups = "drop") %>% 
  ggplot(mapping = aes(x = age, 
                       y = median, 
                       color = INC_Q,
                       fill = INC_Q,
                       ymin = lower,
                       ymax = upper)) + 
  geom_ribbon(alpha=.2, color = NA) +
  geom_line(size=1.5) + 
  scale_fill_brewer(palette = "Paired") + 
  scale_color_brewer(palette = "Paired") +
  xlim(30,78) + 
  ylim(1,6.5) + 
  # to be fixed:
  geom_segment(aes(x=65,y=3.54,xend=65,yend=4.3),color = "black") +
  annotate("text", x = 35, y = 4.5, label = "Women in the lowest\nincome quintile had\non average .8 more\ndisability bouts by age 65",hjust=0) +
  geom_curve(mapping=aes(x=47,y=4.2,xend = 64,yend=4.1),
             color="black",size=.1,curvature=-.3,arrow = arrow(length = unit(0.02, "npc")))+
  annotate("text", x = 46, y = 2.7, label = "lowest 20%", size = 5) +
  annotate("text", x = 50, y = 1.8, label = "highest 20%", size = 5) +
  labs(x = "Age", 
       y = "mean episode order",
       main = "New disability episodes are on average 2nd episodes by age 54 if you're poor, 65 if you're rich")+
  guides(color = "none", fill = "none")+
  theme(
    axis.title.x = element_text(size = 16), # use 16 for presentations
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.y = element_text(size = 16)) +
  theme_minimal()
```

## Inequality in end-of-life disability levels and dispersion
```{r, message=FALSE,warning=FALSE, echo = FALSE}
A1.3 %>% 
  group_by(INC_Q, ad5, ttd) %>% 
  summarize(median = quantile(ttdprev,.5)) %>% 
  ggplot(mapping = aes(x = ad5-ttd, 
                       y = as.factor(ad5), 
                       color = INC_Q,
                       fill = INC_Q,
                       height = median)) + 
  geom_density_ridges(stat = "identity",alpha = .4) +
  scale_fill_manual(values = c("V" = "#3182b0","I" = "#80a8bf")) +
  scale_color_manual(values = c("V" = "#3182b0","I" = "#80a8bf")) +
  annotate("text",x=68,y=5.5,label="highest 20%") + 
  geom_segment(aes(x=63,y=5.5,xend=59,yend=6.1),color = "black") +
  annotate("text",x=39,y=6.8,label="lowest 20%") + 
  geom_segment(aes(x=43.5,y=6.7,xend=48,yend=6.6),color = "black") +
  labs(x = "Age", y = "Prevalence by age at death",size=2) +
  guides(fill="none",color="none") +
  theme(
    axis.title.x = element_text(size = 16),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    legend.position = c(0.9, 0.1)) +
  theme_minimal(  )
```

## Conclusions

\Large
- Help pose and answer questions
- Measure recipes translate to natural language
- Diverse applications
- `R` package `Spells` in beta version

\pause
 ![](SpellsHex.pdf){width=200px}



