---
title: "Alignment, Clocking, and Macro Patterns of Episodes in the Life Course"
author: "Tim Riffe and Andrés Castro"
date: "9 Oct. 2019"
colortheme: "dove"
output: beamer_presentation
header-includes:
  - \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Backstory

\Large
Sequence analysis for *pathways-to-event* questions can be tricky (Hu presentation, 2017) .

\pause

\Large
Idea: Realign sequences on transition. ![](idea.pdf){width=70px}

## Backstory

\Large
Markov matrix expression for average episode count (Dudel, 2017-).

\pause

\Large
Question: What else might we derive?  ![](thought.pdf){width=70px}
 
## Example questions:

\Large
> - Do disability episodes get shorter or longer with age? And over time?
> - What is the distribution of other state episode durations before cancer?
> - How much of an expectancy is composed of short vs long episodes?
> - How do parity-specific birth interval distributions vary by completed fertility?

##  
 
\Large 
\begin{block}{\textbf{Problem}}
Tools for answering such questions are scattered.
\end{block}

<br>
\pause

\begin{block}{\textbf{Corollary}}
Questions are posed less often, and new pattern discovery less frequent.
\end{block}

## 

\Large
\begin{block}{\textbf{Solution}}
We develop a framework (or grammar) of data operations to flexibly derive aggregate patterns.
\end{block}

<br>
\pause

\begin{block}{\textbf{Approach}}
Define episode pre-aggregation operations: \textbf{Clocks}\\
Define time structuring operations: \textbf{Alignment} 
\end{block}
<!-- if simulated then there is likely a markov expression waiting to be written -->

## Approach

\Large
\begin{block}{\textbf{Clocks}}
Within episodes of state \textbf{s}, count time \textcolor{red}{steps} or episode  \textcolor{red}{order} up or down, or total episode  \textcolor{red}{duration} conditional on time of episode entry, exit, or neither.
\end{block}

## Approach

\Large
\begin{block}{\textbf{Alignment}}
  \textcolor{red}{left}, \textcolor{red}{right}, \textcolor{red}{center}, etc. on the \textcolor{blue}{first}, \textcolor{blue}{last}, \textcolor{blue}{longest}, \textcolor{blue}{shortest}, \textcolor{blue}{$n^{th}$}, \textcolor{blue}{$n^{th}$ from last} episode of state \textbf{s}. (and others)
\end{block}

## Illustrations

\Large
10 trajectories simulated from Dudel (2017)
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
cols   <- c("#74ee65", "#773129", "#41bbc5",NA)
states <- c("Employed", "Inactive","Retired","Dead")
yvals  <- 9:0 * 1.1
X      <- readRDS(here::here("Spells","LabTalk","X.rds"))
library(Spells)
source(here::here("Spells","R","DrawSequences.R"))
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Age structured prevalence.

\Large 
\emph{Identity} clock in employment state
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Ones            <- X == "Employed"
Ones[Ones]      <- 1
Ones[Ones == 0] <- NA
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence3(Ones[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Age structured prevalence.

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
XX       <- readRDS(here::here("Spells","LabTalk","RTraj_clean.rds"))
Ones     <- XX == "Employed"
Ones[XX == "Dead"] <- NA
PrevEmpl <- rowMeans(Ones, na.rm=TRUE)
```
```{r}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,1), axes = FALSE, xlab = "", ylab = "")
lines(50:100, PrevEmpl, lwd = 2)
axis(1)
text(43,.5,"Prevalence",xpd=TRUE,srt=90)
axis(2, las = 1)
```

## Illustration: Clocks

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Clocks: Duration (unconditional)

```{r}
# unconditional durations
Dur      <- apply(X,2,clock,clock_type="duration",state = "Inactive")
# conditional on age at entry:
Dur

```
```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```


## Illustration: Clocks: \textcolor{red}{Ascending order}

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
OrdUp   <- apply(X,2,clock,clock_type="order",state = "Employed",increasing=TRUE, step_size = 1)
OrdDown <- apply(X,2,clock,clock_type="order",state = "Employed",increasing=FALSE, step_size = 1)
```
```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence3(OrdUp[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: \textcolor{blue}{Descending order}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence3(OrdDown[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

