---
title: "Alignment, Clocking, and Macro Patterns of Episodes in the Life Course"
author: "Tim Riffe, Angelo Lorenti, Erik Vickstrom, Marcus Ebeling, Andrés Castro"
date: "17 Dec. 2019"
colortheme: "dove"
output: beamer_presentation
header-includes:
  - \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Backstory

\Large
Sequence analysis for *pathways-to-event* questions can be tricky (Yaoyue Hu presentation, 2017) .

\pause

\Large
Idea: Realign sequences on transitions. ![](idea.pdf){width=70px}

## Backstory

\Large
Matrix algebra expression for average episode count (Dudel & Myrskylä, 2017-).

\pause

\Large
![](thought.pdf){width=70px} Question: Are bespoke algebraic derivations necessary? 
 
## Example questions:

\Large
> - Do disability episodes get shorter or longer with age? And over time?
> - What is the distribution of other state episode durations before cancer?
> - How much of an expectancy is composed of short vs long episodes?
> - How do parity-specific birth interval distributions vary by completed fertility or in response to birth outcomes?

##  
 
\Large 
\begin{block}{\textbf{Problem}}
Tools for answering such questions are scattered.
\end{block}

<br>
\pause

\begin{block}{\textbf{Corollary}}
Questions are posed less often, and new pattern discovery less frequent.
\end{block}

## 

\Large
\begin{block}{\textbf{Solution}}
We develop a framework (or grammar) of data operations to flexibly derive aggregate patterns from trajectory data.
\end{block}

<br>
\pause

\begin{block}{\textbf{Approach}}
 \textbf{Clocks} are within and between episode timekeeping operations.\\
 \textbf{Alignment} is a time structuring operation.  
\end{block}
<!-- if simulated then there is likely a markov expression waiting to be written -->

## Approach

\Large
\begin{block}{\textbf{Clocks}}
Within episodes of state \textbf{s}, count time \textcolor{red}{steps} or episode  \textcolor{red}{order} up or down, or total episode  \textcolor{red}{duration} conditional on time of episode entry, exit, or neither.
\end{block}

## Approach

\Large
\begin{block}{\textbf{Alignment}}
  \textcolor{red}{left}, \textcolor{red}{right}, \textcolor{red}{center}, etc. on the \textcolor{blue}{first}, \textcolor{blue}{last}, \textcolor{blue}{longest}, \textcolor{blue}{shortest}, \textcolor{blue}{$n^{th}$}, \textcolor{blue}{$n^{th}$ from last} episode of state \textbf{s}. 
\end{block}

## Requisites

\Large 
\begin{block}{\textbf{Trajectory data}}
 A set of either \textcolor{red}{observed} or \textcolor{red}{simulated} time series of \textcolor{blue}{discrete time steps} consisting in  \textcolor{blue}{categories}. 
\end{block}

\pause
\small
All examples here based on individual multistate (categorical) trajectories in uniform annual time steps

## Illustrations

\Large
10 lives simulated from Dudel & Myrskylä (2017)
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
if (! "Spells" %in% rownames(installed.packages())){
  remotes::install_github("timriffe/Spells/Spells/R/Spells")
}
library(tidyverse)
library(Spells)
library(colorspace)
library(here)
library(reshape2)
library(ggridges)
source(here::here("Spells","R","DrawSequences.R"))
#cols   <- c(qualitative_hcl(5, palette = "Dark 3")[c(3,2,4)],NA)
palette_OkabeIto <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
cols             <- c(palette_OkabeIto[c(3,1,2)],NA)
# plot(NULL, xlim=c(0,length(palette_OkabeIto)), ylim=c(0,1), 
#     xlab="", ylab="", xaxt="n", yaxt="n")
# rect(0:(length(palette_OkabeIto)-1), 0, 1:length(palette_OkabeIto), 1, col=palette_OkabeIto)

# background colors for clock measures:
cols2            <- lighten(desaturate(cols,.3),.3)
states           <- c("Employed", "Inactive","Retired","Dead")
yvals            <- 9:0 * 1.1
X                <- readRDS(here("Spells","CED","X.rds"))
XX               <- readRDS(here("Spells","CED","RTraj_clean.rds"))
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Age structured prevalence.

\Large 
\emph{Identity} clock in employment state
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Ones            <- X == "Employed"
Ones[Ones]      <- 1
Ones[Ones == 0] <- NA
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(Ones[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Age structured prevalence.

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Ones     <- XX == "Employed"
Ones[XX == "Dead"] <- NA
PrevEmpl <- rowMeans(Ones, na.rm=TRUE)
```
```{r}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,1), axes = FALSE, xlab = "", ylab = "")
lines(50:100, PrevEmpl, lwd = 2)
axis(1)
text(43,.5,"Prevalence",xpd=TRUE,srt=90)
axis(2, las = 1)
```

## Illustration: Clocks: Duration (unconditional)

```{r}
# unconditional durations
Dur      <- apply(X,2,clock,clock_type="duration",state = "Inactive")
# conditional on age at entry:
# (function not yet wrapped in clock())
DurEntry      <- apply(X,2,spell_dur_conditional,state = "Inactive", entry = TRUE)
DurExit      <- apply(X,2,spell_dur_conditional,state = "Inactive", entry = FALSE)
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(Dur[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Duration \textcolor{red}{conditioned on entry}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(DurEntry[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Duration \textcolor{blue}{conditioned on exit}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(DurExit[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Order \textcolor{red}{Ascending}

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
OrdUp   <- apply(X,2,clock,clock_type="order",state = "Employed",increasing=TRUE, step_size = 1)
OrdDown <- apply(X,2,clock,clock_type="order",state = "Employed",increasing=FALSE, step_size = 1)
```
```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(OrdUp[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Order \textcolor{blue}{Descending}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(OrdDown[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Steps \textcolor{red}{Ascending}

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
StepUp   <- apply(X,2,clock,clock_type="step",state = "Inactive",increasing=TRUE, step_size = 1)
StepDown <- apply(X,2,clock,clock_type="step",state = "Inactive",increasing=FALSE, step_size = 1)
```

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(floor(StepUp)[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Clocks: Steps \textcolor{blue}{Descending}

```{r,fig.height=4,fig.width=9}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,101), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
  draw_sequence2(X[,i],states,cols2,y=yvals[i], border = NA)
	draw_sequence3(floor(StepDown)[,i],y=yvals[i])
	rect(50,yvals[i],(which(X[,i] == "Dead")[1]+49),yvals[i]+1,border = gray(.4),lwd=.5)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: Age = Birth alignment

```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(50,130), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence2(X[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(50,yvals+.5,1:10,pos=2,xpd=TRUE)
text(47,6,"Random individual i",xpd=TRUE,srt=90)
legend(60,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Alignment: Death

```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
XdeathAlign <- apply(X,2,align,state = "Dead",type="left")
```

```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-80,0), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=XdeathAlign[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-80,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-83,6,"Random individual i",xpd=TRUE,srt=90)
legend(-40,-2,fill = cols, legend = states[-4],horiz = TRUE,xpd=TRUE,bty="n")
```

## Illustration: Alignment: \emph{Entry} to \emph{first} retirement
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
XretirefirstAlign <- apply(X,2,align,state = "Retired",type="left",spell = "first")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=XretirefirstAlign[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-33,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: \emph{Exit} from \emph{first} employment
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Xemployexitfirst <- apply(X,2,align,state = "Employed",type="right",spell = "first")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=Xemployexitfirst[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-33,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: \emph{Exit} from \emph{longest} employment
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Xemployexitlongest<- apply(X,2,align,state = "Employed",type="right",spell = "longest")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=Xemployexitlongest[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-33,6,"Random individual i",xpd=TRUE,srt=90)
```

## Illustration: Alignment: \emph{Centered} on \emph{longest} inactivity
```{r, echo = FALSE, results =" hide", message = FALSE, warning = FALSE}
Xemploycenterlongest<- apply(X,2,align,state = "Inactive",type="center",spell = "longest")
```
```{r,fig.height=4,fig.width=10}
par(mai=c(.8,1,0,0))
plot(NULL, type = "n", xlim = c(-30,50), ylim = c(0,12), axes = FALSE, xlab = "", ylab = "")
for (i in 1:10){
	draw_sequence4(state_seq = X[,i],x=Xemploycenterlongest[,i],states,cols,y=yvals[i], border = NA)
}
axis(1)
text(-30,yvals+.5,1:10,pos=2,xpd=TRUE)
text(-30-3,6,"Random individual i",xpd=TRUE,srt=90)
```

## Aggregation

\Large
\begin{block}{\textbf{Macro patterns}}
Combine clocks and alignment to aggregate (e.g. means, quantiles)
\end{block}

## Application 1: Health

\Large
> - Italian SILC data
> - Ages 20-80
> - Income quintiles
> - Multistate model of disability
> - \textbf{Simulate} discrete life trajectories

## Application 1: Health
```{r statespace, echo = FALSE,warning=FALSE,message=FALSE}
# tools::texi2dvi('statespace.tex',pdf=TRUE,clean=TRUE)
# a<-file.rename(from="statespace.pdf",to="statespace.pdf")
# system(paste("cd",here::here(), "&& inkscape -l statespace.svg statespace.pdf",sep=" "))
knitr::include_graphics("statespace.pdf")
```

```{r, results = 'hide', echo = FALSE, warning=FALSE, message=FALSE}
Dat <- readRDS(here::here("Spells","Data","Lorenti","SILCsim.rds"))

DisStats <- 
  Dat %>% 
  # remove age 80, since we closed out
  filter(age < 80) %>% 
  group_by(InQ, id) %>% 
  # say we want average duration of disability
  # * spells starting in age x
  # * spells ending in age x
  # avg duration of nth spells by age.
  mutate(dis_dur = clock(state, 
                         state = "Disabled",
                         clock_type = "duration"),
         dis_order = clock(state, 
                           state = "Disabled",
                           clock_type = "order",
                           increasing = TRUE)) %>% 
  group_by(InQ, id, dis_order) %>% 
  mutate(first = row_number() == 1L,
         last = row_number() == n(),
         first = ifelse(state == "Disabled" & 
                          max(age) == 79, FALSE, first),
         last = ifelse(state == "Disabled" & 
                          max(age) == 79, FALSE, last)) 
```

## Inequality in disability spell duration
```{r, message=FALSE,warning=FALSE, echo = FALSE}
DisStats %>% 
  filter(first,
         !is.na(dis_dur)) %>% 
  group_by(sex, InQ, age) %>% 
  summarize(dur_first_mean = mean(dis_dur, na.rm = TRUE)) %>% 
  mutate(dur_first_mean = na_if(dur_first_mean, NaN)) %>% 
  ggplot(mapping = aes(x = age, 
                       y = dur_first_mean, 
                       color = InQ)) + 
  geom_line(size=2) + 
  xlim(16,70) +
  labs(x = "Age", y = "mean spell duration",size=2)+
  guides(color=guide_legend(title="Inc. Quintile"))+
  theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 16))
```

## Inequality in end-of-life disability levels and dispersion
```{r, message=FALSE,warning=FALSE, echo = FALSE}
Dat %>%
  filter(age < 80) %>% 
  group_by(InQ, id) %>% 
  mutate(dead = ifelse(any(state == "Dead"),TRUE,FALSE)) %>% 
  ungroup() %>% 
  filter(dead,
         state != "Dead") %>% 
  group_by(InQ, id) %>% 
  mutate(ttd = max(age) - age,
         ad5 = max(age) - max(age) %% 5) %>% 
  ungroup() %>% 
  filter(ad5 > 30) %>% 
  group_by(sex, InQ, ttd, ad5) %>% 
  summarize(ttdprev = mean(state == "Disabled")) %>% 
  ggplot(mapping = aes(x = ad5-ttd, 
                       y = as.factor(ad5), 
                       color = InQ,
                       fill = InQ,
                       height = ttdprev)) + 
  geom_density_ridges(stat = "identity",alpha = .4) +
   labs(x = "Time to death", y = "Prevalence by age at death",size=2)+
  guides(color=guide_legend(title="Inc. Quintile"))+
  theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 16))
```

## How many times have people been disabled? 

```{r, echo = FALSE, warning = FALSE, message = FALSE}
DisStats %>% 
  filter(first,
         !is.na(dis_dur)) %>% 
  group_by(sex, InQ, age) %>% 
  summarize(order_first_mean = mean(dis_order, na.rm = TRUE)) %>% 
  mutate(order_first_mean = na_if(order_first_mean, NaN)) %>% 
  ggplot(mapping = aes(x = age, 
                       y = order_first_mean, 
                       color = InQ)) + 
  geom_line(size = 2) + 
  xlim(16,70) + 
  labs(x = "Age", 
       y = "mean spell order")+
  guides(color=guide_legend(title="Inc. Quintile"))+
  theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 16))
```

## Application 2: Fertility

\Large
> - Colombian DHS data, all waves
> - Birth and union histories
> - Completed fertility >= 2
> - Explore birth intervals
> - Combine clocks and alignment

## Conditional mean time to second birth

```{r, echo = FALSE, warning = FALSE, message = FALSE}
db_tidy <- readRDS(here::here("Spells","Data","Castro","cas_wom_measures.rds"))
db_tidy %>% 
  # aligned on ev union, so need to filter on it.
  group_by(sexf, afb5, left_par1) %>% 
  summarize(med_time_left = median(c_step_down_par1, na.rm=TRUE),
            mean_time_left = mean(c_step_down_par1, na.rm=TRUE))  %>% 
  ggplot(mapping = aes(x = left_par1, 
                       y = mean_time_left, 
                       color = as.factor(sexf))) + 
  geom_line() + 
  #geom_vline(xintercept = c(0,1)) +
  # transparent band presumed to isolate twins
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 5,
           alpha = .2) +
  facet_wrap(~afb5) + 
  xlim(0,15) + 
  guides(color=guide_legend(title="Sex of first child")) + 
  xlab("time since first birth") + 
  ylab("mean time to second birth")+
   theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 16))
```

## Does a first boy imply a faster next boy?

```{r, echo = FALSE, warning = FALSE, message = FALSE}
db_tidy %>% 
  # aligned on ev union, so need to filter on it.
  group_by(sexf, afb5, left_par1) %>% 
  summarize(med_time_left = median(c_step_down_boy1, na.rm=TRUE),
            mean_time_left = mean(c_step_down_boy1, na.rm=TRUE))  %>% 
  ggplot(mapping = aes(x = left_par1, 
                       y = mean_time_left, 
                       color = as.factor(sexf))) + 
  geom_line() + 
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 8,
           alpha = .2) +
  facet_wrap(~afb5) + 
  guides(color=guide_legend(title="Sex of first child")) + 
  xlab("time since first birth") + 
  ylab("mean time to next boy") +
  xlim(0,15)+
     theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 16))
```

## Does a first girl imply a faster next girl?

```{r, echo = FALSE, warning = FALSE, message = FALSE}
db_tidy %>% 
  # aligned on ev union, so need to filter on it.
  group_by(sexf, afb5, left_par1) %>% 
  summarize(med_time_left = median(c_step_down_gir1, na.rm=TRUE),
            mean_time_left = mean(c_step_down_gir1, na.rm=TRUE))  %>% 
  ggplot(mapping = aes(x = left_par1, 
                       y = mean_time_left, 
                       color = as.factor(sexf))) + 
  geom_line() + 
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 8,
           alpha = .2) +
  facet_wrap(~afb5) + 
  guides(color=guide_legend(title="Sex of first child")) + 
  xlab("time since first birth") + 
  ylab("mean time to next boy") +
  xlim(0,15)+
     theme(
  axis.title.x = element_text(size = 16),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 16))
```

## Conclusions

\Large
> - Help pose and answer questions
> - Measures translate to natural language
> - `R` package `Spells` in beta version
> - Grammar still in development
> - Searching for diverse applications

\pause
 ![](SpellsHex.pdf){width=200px}



